### 2018-03-17 paper/2-background.tex
Proof S2

If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
will first build \cmd{util.o} and \cmd{main.o}, in any order since the
corresponding tasks are independent, and then build \cmd{main.exe}. If the
user modifies the sources of \cmd{util.h} and runs \Make again, it will
perform a \emph{full rebuild}, because all three tasks transitively depend on
\cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if
the user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
\emph{acyclic} then each task needs to be executed at most once. Cyclic task
dependencies are typically not allowed in build systems although there are rare
exceptions, see~\S\ref{sec-iterative-compute}.

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
not known statically\footnote{One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}.

\Excel handles this example correctly: if \cmd{C1=1} and the user changes
\cmd{A2}, \Excel will not recompute the result \cmd{B1}. Alas, other forms of
dynamic dependencies can force \Excel to perform unnecessary computation.
Consider the following modification of the above example:

@@

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This clearly violates the minimality
property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
will recompute \cmd{B1}, potentially triggering further unnecessary
recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel can
also track changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single build task has changed. We further discuss
self-tracking in~\S\ref{sec-tracking-aspects}.

@@

When build systems are used by large teams, different team members often end up
executing exactly the same tasks on their local machines. A \emph{cloud build
system} can speed up builds dramatically by transparently sharing build results
among team members. Furthermore, cloud build systems allow one to perform
\emph{shallow builds} that materialise only end build products locally, leaving
all intermediates in the cloud. We illustrate shallow cloud builds by an example
(see~Fig.~\ref{fig-bazel}).

@@

In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
source \cmd{util.c}, thereby changing its hash from~1~to~7. The build system
finds that nobody has ever compiled the new $\{\cmd{util.c}, \cmd{util.h}\}$
combination and must therefore build \cmd{util.o}, which results in changing its
hash from~4~to~8. The obtained combination of hashes of \cmd{util.o} and
\cmd{main.o} has not been encountered before either, therefore the build system
first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
linking the two object files. When the build is complete, the obtained results
may be uploaded to the cloud for future reuse by other team members.

\Bazel is one of the first examples of openly available cloud build systems.
Like \Make, it does not support dynamic dependencies and can therefore benefit
from the simplicity of building tasks in a statically known topological order.
It is minimal and supports the early cutoff optimisation. To support cloud
builds, \Bazel maintains a \emph{content-addressable cache} that can be used to
fetch a previously built file given the hash of its content, and
\emph{dependency graphs from all previous builds}, annotated with observed file
hashes. The latter allows to bypass the execution of a task, by predicting the
hash of the result from the hashes of its dependencies, and subsequently
fetch the result from the cache. A concrete implementation is provided
in~\S\ref{sec-implementations}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that. In principle a single \emph{dirty bit} per file is
    sufficient, as will be demonstrated in~\S\ref{sec-implementations}.
    \item \Excel stores one dirty bit per cell and the calculation chain from
    the previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

@@

In this paper we elucidate which build system properties are consequences of
specific implementation choices (metadata and algorithm), and how one can obtain
new build systems with desired properties by recombining parts of existing
implementations. As a compelling example, we demonstrate how to combine
advantages of \Shake and \Bazel in a cloud build system with dynamic
dependencies, see~\S\ref{sec-implementation-cloud-shake}.

>>>>

If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
will first build \cmd{util.o} and \cmd{main.o}, in any order since these
tasks are independent, and then build \cmd{main.exe}. If the
user modifies the sources of \cmd{util.h} and runs \Make again, it will
perform a \emph{full rebuild}, because all three tasks transitively depend on
\cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if
the user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
\emph{acyclic} then each task needs to be executed at most once. Cyclic task
dependencies are typically not allowed in build systems although there are rare
exceptions, see~\S\ref{sec-iterative-compute}.

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, which is
not known statically\footnote{One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}.

\Excel handles this example correctly: if \cmd{C1=1} and the user changes
\cmd{A2}, \Excel will not require \cmd{B1} to be computed in advance. Alas, other forms of
dynamic dependencies can force \Excel to perform unnecessary computation.
Consider the following modification of the above example:

@@

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This approach clearly violates the minimality
property~\ref{def-minimal}: if \cmd{C1=1} and the user modifies \cmd{A2}, \Excel
will recompute \cmd{B1}, potentially triggering further unnecessary
recalculation even though \cmd{B1} does not transitively depend on \cmd{A2}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel can
also track changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single build task has changed. We discuss
self-tracking further in~\S\ref{sec-tracking-aspects}.

@@

When build systems are used by large teams, different team members often end up
executing exactly the same tasks on their local machines. A \emph{cloud build
system} can speed up builds dramatically by sharing build results
among team members. Furthermore, cloud build systems allow one to perform
\emph{shallow builds} that materialise only end build products locally, leaving
all intermediates in the cloud. We illustrate shallow cloud builds by an example
in Fig.~\ref{fig-bazel}.

@@

In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
source \cmd{util.c}, thereby changing its hash from~1~to~7. The build system
finds that nobody has ever compiled the new $\{\cmd{util.c}, \cmd{util.h}\}$
combination and must therefore build \cmd{util.o}, which results in changing its
hash from~4~to~8. The combination of hashes of \cmd{util.o} and
\cmd{main.o} has not been encountered before either, therefore the build system
first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
linking the two object files. When the build is complete, the results
can be uploaded to the cloud for future reuse by other team members.

\Bazel is one of the first examples of openly available cloud build systems.
Like \Make, it does not support dynamic dependencies and can therefore benefit
from the simplicity of building tasks in a statically known topological order.
It is minimal and supports the early cutoff optimisation. To support cloud
builds, \Bazel maintains a \emph{content-addressable cache} that can be used to
fetch a previously built file given the hash of its content, and
\emph{dependency graphs from all previous builds}, annotated with observed file
hashes. The latter allows builds to bypass the execution of a task, by predicting the
hash of the result from the hashes of its dependencies, and subsequently
fetch the result from the cache. A concrete implementation is provided
in~\S\ref{sec-implementations}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that.
    \item \Excel stores one dirty bit per cell and the calculation chain from
    the previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

@@

In this paper we elucidate which build system properties are consequences of
specific implementation choices (metadata and algorithm), and how one can obtain
new build systems with desired properties by recombining parts of existing
implementations. As a compelling example, we demonstrate how to combine the
advantages of \Shake and \Bazel in a cloud build system with dynamic
dependencies, see~\S\ref{sec-implementation-cloud-shake}.
